Sequential rank agreement methods for comparison of ranked lists

### 2016-06-14 latex/superranker-manuscript.tex
Part I of corrections

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and prominent task in applied statistics. Predictor
  variables are ranked according to magnitude of association with an
  outcome, prediction models rank subjects according to the
  personalized risk of an event, and genetic studies rank genes
  according to their difference in gene expression levels. This
  article constructs measures of the agreement of two or more ordered
  lists. We use the standard deviation of the ranks to define a
  measure of agreement that both provides an intuitive interpretation
  and can be applied to any number of lists even if some or all are
  incomplete or censored. The approach can identify change-points in
  the agreement of the lists and the sequential changes of agreement
  as a function of the depth of the lists can be compared graphically
  to a permutation based reference set.  The usefulness of these tools
  are illustrated using gene rankings, and using data from two Danish
  ovarian cancer studies where we assess the within and between
  agreement of different statistical classification methods.
\end{abstract}

@@

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and the provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

Other recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. One proposal is denoted the average
overlap (AO) and used for comparison later on. Specifically,
\citet{Webber2010} define their rank-biased overlap (RBO) by weighting
with a converging series to ensure that the top is weighted higher
than the potentially non-informative bottom of the lists.
It is possible to use the existing methods to calculate agreement of
lists until a given depth, i.e., limited to the $d$ items of each
list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

Very recently, \citet{hall:schi:2012} proposed a method for comparing
pairwise rankings and derived the asymptotic distribution of the
endpoint where the two ranked lists no longer are in agreement. Their
approach was based on anchoring one of the two lists and subsequently
generating a sequence of 0s and 1s depending on whether the ranks in
the second list was close to the rank from the anchored
list. \citet{samp:verd:2013} followed up on this idea for pairwise
comparison of lists but used penalties based on a truncated geometric
probability instead of a 0-1 process and they evaluated the
distribution of the endpoint of agreement by computational approaches.
The asymptotic distribution in the \citet{hall:schi:2012} paper is
based on letting the number of \emph{lists} increase to infinity which
is a situation that is only relevant in special cases, and the
simulation-based null distribution approach of \citet{samp:verd:2013}
is guaranteed to work also for small samples.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could be essentially be used. Our proposed approach allows us to
compare more than two lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, places
high weight on items at the top of the list, accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
change-point, i.e., a list depth where a change in the agreement of
the lists occur but we also provide randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored/partial lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker}. \mcomment{XXX check dette
passer, submit pakke}

% chunk 1

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$. In all what follows we condition on the set of
items and consider the ranking function to be a random variable. Thus,
let $R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $(P!)^{-1}$ to each of
the $P!$ different rankings.
Panels (a) and (b) of
Table~\ref{tab:example} show a schematic example of these
mappings. Thus if $R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked
first in list $l$ and similarly $R_l(X_{34})=1$.

\bigskip\hrule HOV: er det virkelig equivalent ogsaa mht vores nye theoretiske
ideer? Og bruger vi det nogen steder?  enten skal det forklares med et
konkret eksempel eller ud?  We will use the same notation for the
equivalent scenario where we have ordered lists obtained from applying
the same ranking function to $L$ samples from the same population,
i.e., for ranked lists $R(X^1), \ldots, R(X^L)$.  
\hrule\bigskip

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked list of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_P)-\E_Q R(X_P)\right)^2\right]\\
   &= \sum_{r=1}^{P!} \left(r(X_P)-\E_Q R(X_P)\right)^2 Q(r)\nn
\end{align}
where $\E_Q R(X_p)=\sum_{r=1}^{P!} r(X_p)Q(r)$ and $Q$ is a probability mass function
on the space of all permutations of lists with $P$ elements.
The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p)\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r=1}^{P!} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.
We then define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{align}
  \textrm{sra}(d) = \frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)\label{def:sra}
\end{align}
where $|S(d)|$ is the cardinality of the set $S(d)$.
The empirical counterpart is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero suggests that the lists agree
while larger values suggest disagreement.
Just like in method comparison studies we prefer smaller limits of
agreement because that means that the differences between the methods
is small. If the ranked lists are identical then the value of
sequential rank agreement will be zero for all depths $d$.
The sequential rank agreement is  equivalent to the pooled
variance of the items found in $S(d)$, and it can be interpreted as the
average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.
For an easier interpretation we suggest to plot $\widehat{\textrm{sra}}_L(d)^{1/2}$ 
corresponding to the pooled standard deviation.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify change-points in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and they both provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

Other more recent approaches consider the intersection of lists as the basis
for a similarity measure. However, simple intersection also places
equal weights on all depths of the list and therefore \citet{Fagin2003} and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists. 
% One proposal is denoted the average overlap (AO) and used for
% comparison later on. 
Specifically, \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series to ensure that the top is
weighted higher than the potentially non-informative bottom of the
lists.  It is possible to use the existing methods to calculate
agreement of lists until a given depth, i.e., limited to the $d$ items
of each list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

Very recently, \citet{hall:schi:2012} proposed a method for comparing
pairwise rankings and derived the asymptotic distribution of the
endpoint where the two ranked lists no longer are in agreement. Their
approach was based on anchoring one of the two lists and subsequently
generating a sequence of 0s and 1s depending on whether the ranks in
the second list was close to the rank from the anchored
list. \citet{samp:verd:2013} followed up on this idea for pairwise
comparison of lists but used penalties based on a truncated geometric
probability instead of a 0-1 process and they evaluated the
distribution of the endpoint of agreement by computational approaches.
The asymptotic distribution in the \citet{hall:schi:2012} paper is
based on letting the number of \emph{lists} increase to infinity which
is a situation that is only relevant in special cases, whereas the
simulation-based null distribution approach of \citet{samp:verd:2013}
is guaranteed to work also for small samples.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could be essentially be used. Our proposed approach allows us to
compare more than two lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, places
high weight on items at the top of the list, accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
change-point, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete/censored/partial lists. In section 3 we present and
discuss approaches to evaluate the results obtained from sequential
rank agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies before we discuss the findings along
with possible extensions. The approaches presented in this manuscript
are available in the R package \texttt{SuperRanker} which can be found
on CRAN.
% chunk 1

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$.  Panels (a) and (b) of Table~\ref{tab:example}
show a schematic example of these mappings. Thus if
$R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked first in list $l$
and similarly $R_l(X_{34})=1$.

In all what follows we consider a fixed set of items and consider the
ranking function to be a random variable. Thus, let
$R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $1/P!$ to each of
the $P!$ different possible rankings.

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked lists of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

The agreement of the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_P)-\E_Q R(X_P)\right)^2\right]\\
   &= \sum_{r=1}^{P!} \left(r(X_P)-\E_Q R(X_P)\right)^2 Q(r),\nn
\end{align}
\textsf{Ikke så vild med notationen her ovenfor}
where $\E_Q R(X_p)=\sum_{r=1}^{P!} r(X_p)Q(r)$ and $Q$ is a probability mass function
on the space of all permutations of lists with $P$ elements.
The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r=1}^{P!} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{align}
  \textrm{sra}(d) = \frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)\label{def:sra}
\end{align}
where $|S(d)|$ is the cardinality of the set $S(d)$.
The empirical counterpart is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the methods is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all depths $d$.  The sequential rank
agreement is equivalent to the pooled variance of the items found in
$S(d)$, and it can be interpreted as the average distance of the
individual rankings of the lists from the average ranking for each of
the items we have seen until depth $d$.  For an easier interpretation
we suggest to plot $\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding
to the pooled standard deviation.
